Rapid-response Bayesian sky localization for electromagnetic follow-up of gravitational-wave candidates

### 2013-07-26 ms.tex
'Precalculated' is not a word

The innermost integration is over $D_\mathrm{L}$ and is performed using adaptive Gaussian quadrature, but with a twist that speeds up convergence. As in many parameter estimation problems, the integrand may be sharply peaked, and any generic adaptive integration scheme may take a long time to find the peak. Before evaluating each inner integral over $D_\mathrm{L}$, the quantities $A=-\frac{1}{2} \sum_i {\rho_i}^2$ and $B=\sum_i \rho_i \hat\rho_i$ are precalculated so that the integrand can be written as
%
\begin{equation*}
	\exp\left(\frac{A}{{D_\mathrm{L}}^2} + \frac{B}{D_\mathrm{L}}\right) {D_\mathrm{L}}^m.
\end{equation*}
%
Then, we see that the likelihood $\exp(A {D_\mathrm{L}}^{-2} + B {D_\mathrm{L}}^{-1})$ is maximized when $D_\mathrm{L} = D_{\mathrm{L},0} = -2A/B$. The likelihood takes on a factor $\eta$ (say, $\eta=0.01$) of its maximum value when
%
\begin{equation}
	D_\mathrm{L} = D_{\mathrm{L},\pm} = \left(\frac{1}{D_{\mathrm{L},0}} \mp \sqrt{\frac{\log\eta}{A}}\right)^{-1}.
\end{equation}
%
We have now identified up to five breakpoints that partition the distance integrand into up to four intervals with quantitatively distinct behavior. These intervals are depicted in Figure~\ref{fig:radial_integrand} with distance increasing from left to right. There is a left\nobreakdashes-hand or small distance tail in which the integrand is small and monotically increasing, a left\nobreakdashes- and right\nobreakdashes-hand side of the maximum likelihood peak, and a right\nobreakdashes-hand tail in which the integrand is small and monotonically decreasing. These breakpoints are:
%
\begin{equation}
    D_{\mathrm{L},\mathrm{break}} = \{ D_\mathrm{L} \in
    \left\{
    \begin{array}{c}
    D_{\mathrm{L},\mathrm{min}} \\
    D_{\mathrm{L},-} \\
    D_{\mathrm{L},0} \\
    D_{\mathrm{L},+} \\
    D_{\mathrm{L},\mathrm{max}}
    \end{array}
    \right\} :
    D_{\mathrm{L},\mathrm{min}} \leq D_\mathrm{L}
    \leq D_{\mathrm{L},\mathrm{max}}\}.
\end{equation}
%
We use these breakpoints as initial subdivision in an adaptive Gaussian quadrature algorithm. Specifically, we use the GNU Scientific Library's \verb|gsl_integrate_qagp|\footnote{\url{http://www.gnu.org/software/gsl/manual/html_node/QAGP-adaptive-integration-with-known-singular-points.html}} function, which is designed to integrate functions with known singular points (though, of course, or integrand has no singular points). This function estimates the integral over each subdivision and each interval's contribution to the total error, then subdivides the interval whose error contribution is largest. Subdivisions continue until a fixed total fractional error is reached. In this way, most integrand evaluations are expended on the most important distance interval, whether that happens to be the tails (when the posterior is dominated by the prior) or the peak (when the posterior is dominated by the observations).

>>>>

The innermost integration is over $D_\mathrm{L}$ and is performed using adaptive Gaussian quadrature, but with a twist that speeds up convergence. As in many parameter estimation problems, the integrand may be sharply peaked, and any generic adaptive integration scheme may take a long time to find the peak. Before evaluating each inner integral over $D_\mathrm{L}$, the quantities $A=-\frac{1}{2} \sum_i {\rho_i}^2$ and $B=\sum_i \rho_i \hat\rho_i$ are pre\nobreakdashes-calculated so that the integrand can be written as
%
\begin{equation*}
	\exp\left(\frac{A}{{D_\mathrm{L}}^2} + \frac{B}{D_\mathrm{L}}\right) {D_\mathrm{L}}^m.
\end{equation*}
%
Then, we see that the likelihood $\exp(A {D_\mathrm{L}}^{-2} + B {D_\mathrm{L}}^{-1})$ is maximized when $D_\mathrm{L} = D_{\mathrm{L},0} = -2A/B$. The likelihood takes on a factor $\eta$ (say, $\eta=0.01$) of its maximum value when
%
\begin{equation}
	D_\mathrm{L} = D_{\mathrm{L},\pm} = \left(\frac{1}{D_{\mathrm{L},0}} \mp \sqrt{\frac{\log\eta}{A}}\right)^{-1}.
\end{equation}
%
We have now identified up to five breakpoints that partition the distance integrand into up to four intervals with quantitatively distinct behavior. These intervals are depicted in Figure~\ref{fig:radial_integrand} with distance increasing from left to right. There is a left\nobreakdashes-hand or small distance tail in which the integrand is small and monotically increasing, a left\nobreakdashes- and right\nobreakdashes-hand side of the maximum likelihood peak, and a right\nobreakdashes-hand tail in which the integrand is small and monotonically decreasing. These breakpoints are:
%
\begin{equation}
    D_{\mathrm{L},\mathrm{break}} = \{ D_\mathrm{L} \in
    \left\{
    \begin{array}{c}
    D_{\mathrm{L},\mathrm{min}} \\
    D_{\mathrm{L},-} \\
    D_{\mathrm{L},0} \\
    D_{\mathrm{L},+} \\
    D_{\mathrm{L},\mathrm{max}}
    \end{array}
    \right\} :
    D_{\mathrm{L},\mathrm{min}} \leq D_\mathrm{L}
    \leq D_{\mathrm{L},\mathrm{max}}\}.
\end{equation}
%
We use these breakpoints as initial subdivision in an adaptive Gaussian quadrature algorithm. Specifically, we use the GNU Scientific Library's \verb|gsl_integrate_qagp|\footnote{\url{http://www.gnu.org/software/gsl/manual/html_node/QAGP-adaptive-integration-with-known-singular-points.html}} function, which is designed to integrate functions with known singular points (though, of course, or integrand has no singular points). This function estimates the integral over each subdivision and each interval's contribution to the total error, then subdivides the interval whose error contribution is largest. Subdivisions continue until a fixed total fractional error is reached. In this way, most integrand evaluations are expended on the most important distance interval, whether that happens to be the tails (when the posterior is dominated by the prior) or the peak (when the posterior is dominated by the observations).
